Simulated Observations of Young Protoplanetary Discs

### 2013-02-20 paper.tex
included PC 4th set of changes

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{Tom Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
Paola Caselli$^{1}$,
Aaron Boley$^{2}$,
Richard Durisen$^{3}$,
Tom Hartquist$^{1}$,
John Ilee$^{1}$ and
Jonathan Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\}}

@@

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (J\o rgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimetre and millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). Zapata et al. (2013) used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA will be completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \newline

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission and measure the structure and possible fragmentation (Cossins et al. 2010; Ruge et al. 2013). Molecular line emission from massive discs has been simulated by Krumholz et al. (2007), assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., Visser et al. (2009, 2011). As shown by Ilee et al. (2011; hereafter I2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disc  within a dense core with characteristics similar to the well-studied pre-stellar core L1544, (Keto \& Caselli 2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere (Bonnor 1956; Ebert 1957) with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Caselli et al. 2012; Keto et al. in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}. \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small.eps}
 \caption{The spherically symmetric model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing temperature (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity / 100 (cyan) in m$\,$s$^{-1}$. Adapted from KC2010 and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT6.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, highlighting the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.eps}
 \caption{Column densities of OCS, H$_2$CO,HCO$^+$ and CO used in the disc model. Figure adapted from I2011}
 \label{Chemistry} 
\end{figure*}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane with best fit curve}
 \label{velocity}
\end{figure}

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the young protoplanetary disc has been plugged in. The disc structure is derived by the hydrodynamic model of Boley (2007) and Boley (2009). The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating disc featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout both sections model and the opacities were adopted from Ossenkops and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}, the abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM, Wilson \& Rood 1994). As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar objects. The molecular data used is from the lamda database (Sch\"oier et al. (2005) http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used is LIME (Brinch \& Hogerheijde 2010),which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the grid-construction section (section \ref{gridding}). At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/grid.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the griding process and the paths down which photons can propagate overlaid on the density distribution (in m$^{-3}$, as used in LIME). The points are more concentrated at small radii and in the densest regions.}
 \label{grid}
\end{figure}

@@

We considered C$^{17}$O 3-2 at 337.1GHz with an upper energy level of $E_u$=32.3K, HCO$^+$ 3-2 at 267GHZ and $E_u$=25.7K, OCS 26-25 at 316GHz and $E_u$=205K and H$_2$CO $4_{04}$-3$_{03}$ at 290GHz and $E_u$=34.9K. Other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and CS.\newline

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and the disc is inclined at 30$^\circ$ to edge on. From these simulated observations, integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity.
(Note the integrated intensity and intensity weighted velocity maps (figures \ref{sim_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams  (figure \ref{pvs}) as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions})\newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum.eps}
 \caption{{\bf Left:}A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite griding.}
 \label{averages}
\end{figure}

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all.eps}
 \caption{{\bf Left:}Continuum subtracted integrated intensity maps, {\bf Right:} Intensity weighted velocity maps. The spiral structure in different regions of the disc is highlighted by looking at different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are integrated over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

@@

The abundance of the H$_2$CO molecule is typical of the majority of the molecules simulated in I2011 in that it traces the spiral structure of the inner $\sim$40$\,$AU of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation of detected out to larger radii than is possible with the OCS (figure \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}$-3$_{03}$ line is seen in absorption against the disc mid-plane continuum with the central region seen in emission. Unlike the OCS however, the H$_2$CO extends out to large enough radii to be present in the voids between spiral arms. In these regions the H$_2$CO line can be seen in emission, this can be seen most clearly in the H$_2$CO channel map (figure \ref{h2co_chanmap}).\newline

The fractional abundance of C$^{17}$O is constant at 2$\times$10$^{-8}$ across the disc, meaning the C$^{17}$O line is the most accurate in reproducing the physical structure of the disc. Like H$_2$CO, C$^{17}$O shows emission in voids between spiral arms. The C$^{17}$O line is visible from inner edge of the disc to its outer edge. Like OCS, C$^{17}$O is not seen in outflows (e.g. Yildiz et al. 2012) and so rotation seen in C$^{17}$O lines should only be due to rotation from the disc and not contamination from outflows.\newline

HCO$^+$ in unique in the molecules simulated in I2011 in that it traces only the outer regions of the disc, and so can be used to look at the extended velocity and physical structure. In figure \ref{sim_all} we can see that the HCO$^+$ line shows the outer edges of the spiral arms in absorption, but also shows emission from the diffuse gas further out in the disc. This allows the velocity structure of the disc measured out to larger radii. This allows constraints to be placed on the rotation of the disc out to larger radii than other molecules.

\section{ALMA Predictions} \label{sec:alma_predictions}

In order to make predictions on the observability of the synthetic brightness maps CASA (common astronomy software applications) was used to simulate their observation with the completed ALMA in configuration 26 of 28, giving a beam size of between 0.02 and 0.03 arcseconds. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This requires around 6 hours of integration time. The exact sensitivities are give in table \ref{sigmas}.
\begin{table}
 \centering
 \begin{minipage}{80mm}
   \caption{Sensitivities of ALMA given 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
   \label{sigmas}
   \begin{tabular}{c||c|c|c}
     \hline
     Species & Transition & Frequency & Sensitivity\\
     \hline
     C$^{17}$O & J=3$\rightarrow$2 & 337.06GHz & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
     OCS & J=26$\rightarrow$25 & 316.15GHz & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
     HCO$^+$ & J=3$\rightarrow$2 & 267.56GHz & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
     H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ & 290.62GHz & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
     \hline
   \end{tabular}
 \end{minipage}
\end{table}
In order to clean the images which featured mainly absorption the sky model used was the output of the lime simulations with the continuum subtracted and then the image inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ given in table \ref{sigmas}.\newline

@@

 \caption{Continuum emission at 337GHz simulated for ALMA most extended configuration}
 \label{continuum}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/channel_map-1.eps} 
 \caption{H$_2$CO 4$_{04}$-3$_{03}$ channel maps as simulated by CASA. H$_2$CO line in contours starting at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. Each map integrates over 2.4 km$\,$s$^{-1}$ with the top left from -10.8 to -8.4 km$\,$s$^{-1}$ to the bottom left ranging from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}$-$3_{03}$ line overlaid on the 1$\,$mm continuum emission (figure \ref{continuum}). From this it can be seen that the absorption features trace the disc spiral structure whilst line emission emanates from gaps in the disc where there is less continuum to be absorbed.\newline

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_30deg_contSub_nums.eps}
 \caption{Integrated intensity maps for the simulated observations of:{\bf Top Left:} H$_2$CO 4$_{04}\rightarrow$3$_{03}$, {\bf Top Right:}C$^{17}$O 3$\rightarrow$2, {\bf Bottom Left:} OCS 26$\rightarrow$25 and {\bf Bottom Right:} HCO$^+$ 3$\rightarrow$2. contours start at 3$\sigma$ and increasing in intervals of 3$\sigma$ for each of the maps except the H$_2$CO which starts at 5$\sigma$ and increases in intervals of 5$\sigma$. Emission is in red contours, absorption in blue. The numbers refer to the spectra shown in figure \ref{spectra}.}
\label{mom0_maps}
\end{figure*}

>>>>

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{1}$,
A. C. Boley$^{2}$,
T. W. Hartquist$^{1}$,
R. Durisen$^{3}$, and
J. Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\}}

@@

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources (Andr\'e et al. 2000), finding masses between 0.4 and  $>$1 M$_{\odot}$ (J\o rgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using the Jansky Very Large Array (JVLA) to observe NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). Zapata et al. (2013) used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \newline

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation (Cossins et al. 2010; Ruge et al. 2013). Molecular line emission from massive discs has been simulated by Krumholz et al. (2007), assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., Visser et al. (2009, 2011). As shown by Ilee et al. (2011; hereafter I2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

To describe the full physical enviroment which we expect these young discs to occur in, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (Keto \& Caselli 2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere (Bonnor 1956; Ebert 1957) with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Caselli et al. 2012; Keto et al. in preparation). Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small.eps}
 \caption{The spherically symmetric model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing gas (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity / 10 (cyan) in m$\,$s$^{-1}$. Adapted from KC2010 and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT6.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, highlighting the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.eps}
 \caption{Column densities of OCS, H$_2$CO,HCO$^+$ and CO used in the disc model. Figure adapted from I2011.}
 \label{Chemistry} 
\end{figure*}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane with best fit curve.}
 \label{velocity}
\end{figure}

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of Boley (2007) and Boley (2009). The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The opacities were adopted from Ossenkops and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}, the abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM, Wilson \& Rood 1994). As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. (2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used is LIME (LIne Modeling Engine; Brinch \& Hogerheijde 2010), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in appendix  \ref{sec:gridding}). At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/grid.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate overlaid on the density distribution (in m$^{-3}$, as used in LIME). The points are more concentrated at small radii and in the densest regions.}
 \label{grid}
\end{figure}

@@

We considered C$^{17}$O 3$\rightarrow$2 at 337.1$\,$GHz with an upper energy level of $E_u$=32.3K, HCO$^+$ 3$\rightarrow$2 at 267$\,$GHz and $E_u$=25.7K, OCS 26$\rightarrow$25 at 316$\,$GHz and $E_u$=205K and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ at 290$\,$GHz and $E_u$=34.9K.\newline

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and the disc is inclined at 30$^\circ$ to edge on. From these simulated observations, integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
(The integrated intensity and intensity weighted velocity maps (figures \ref{sim_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams  (figure \ref{pvs}) as the strong absorption feature at all positions around zero velocity. Moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions})\newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum.eps}
 \caption{{\bf Left:}A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite gridding.}
 \label{averages}
\end{figure}

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structure in different regions of the disc is highlighted by looking at different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are integrated over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

@@

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$AU of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation is detected out to larger radii than is possible with the OCS line (figure \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$, is seen in absorption against the disc mid-plane continuum, with the central region seen in emission. Unlike the OCS however, the H$_2$CO extends out to large enough radii to be present in the voids between the outer parts spiral arms. In these regions, the H$_2$CO line can be seen in emission.\newline

The fractional abundance of C$^{17}$O is constant at 2$\times$10$^{-8}$ across the disc, meaning the C$^{17}$O line is the most accurate in reproducing the physical structure of the disc. Like H$_2$CO, C$^{17}$O shows emission in voids between spiral arms. The C$^{17}$O line is visible from the inner edge of the disc to its outer edge. Like OCS, C$^{17}$O is not seen in outflows (e.g. Yildiz et al. 2012) and so rotation seen in C$^{17}$O lines should only be due to rotation from the disc and not contamination from outflows.\newline

HCO$^+$ in unique in the molecules simulated in I2011 in that it traces only the outer regions of the disc, and so can be used to look at the extended velocity and physical structure. In figure \ref{sim_all} we can see that the HCO$^+$ line shows the outer edges of the spiral arms in absorption, but also shows emission from the diffuse gas further out in the disc. This allows the velocity structure of the disc measured out to larger radii. This allows constraints to be placed on the rotation of the disc out to larger radii than with other molecules.

\section{Predictions for ALMA} \label{sec:alma_predictions}

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended comfiguration (out of a total of 28), with a longest baseline of 14.4km and giving beam sizes of between 0.02 and 0.03 arcseconds at the simulated frequencies. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table}
 \centering
 \begin{minipage}{80mm}
   \caption{Sensitivities of ALMA given 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
   \label{sigmas}
   \begin{tabular}{c||c|c|c}
     \hline
     Species & Transition & Frequency & Sensitivity\\
     \hline
     C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
     OCS & 26$\rightarrow$25 & 316.15GHz & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
     HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
     H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ & 290.62GHz & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
     \hline
   \end{tabular}
 \end{minipage}
\end{table}
In order to clean the images, which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the image inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ given in table \ref{sigmas}.\newline

@@

 \caption{CASA simulation on the continuum emission of the model at 337GHz, using the same ALMA configuration as used for the molecular lines.}
 \label{continuum}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/channel_map-1.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated by CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. Each map integrates over 2.4 km$\,$s$^{-1}$ with the top left from -10.8 to -8.4 km$\,$s$^{-1}$ to the bottom left ranging from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission (figure \ref{continuum}). From this it can be seen that the absorption features trace the disc spiral structure whilst line emission emanates from gaps in the outer disc where there is less continuum to be absorbed.\newline
